Income Disparity and Its Influence on Higher Education Affordability and Access

### 2024-02-13 starter_folder-main 4/paper/paper.qmd
updated data cleaning script

1. Recreating the Model: We will start by reproducing the competitive college market model used in the original study, ensuring our work is entirely reproducible. This step includes verifying the model's assumptions, computational methods, and data inputs.
2. Expanding the Study with a Canadian Lens: Specifically examine how income inequality within Canada affects university tuition, comparing these dynamics with the findings from the U.S. context. This will provide insights into how different policy environments and socio-economic conditions influence the relationship between income inequality and higher education costs.

@@

```{r}
#| label: fig-stickers
#| fig-cap: Public vs. Private Sticker and Net Prices Over Time
#| echo: false
#| warning: false

# Load necessary libraries
library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)

# Set the file path
file_path <- "/Users/michaelfang/Downloads/Figure_1_data.xlsx"

# Read the data from the Excel file
data <- read_excel(file_path, sheet = "FIGURE 1 in PAPER", range = cell_cols("D:H"))

# Set column names manually
colnames(data) <- c('Academic Year', 'Public Sticker', 'Public Net', 'Private Sticker', 'Private Net')

# Check the structure of the data to make sure it's what we expect
#str(data)

@@

```{r}
#| echo: false
#| warning: false
#| message: false

@@

# Define the path to your Excel file
file_path <- "/Users/michaelfang/Downloads/SCF_AER_submit.xlsx"

@@

# Plot the data
ggplot(data_long, aes(x = GridLogIncome, y = EMG_ML_PDF, color = Year)) +
  geom_line() +
  labs(title = "Comparison of Income Distribution for 1989 and 2016",
       x = "Log Household Income",
       y = "Density") +
  theme_minimal() +
  scale_color_manual(values = c("2016" = "blue", "1989" = "red"))
```

>>>>

1. Recreating the Model: We will start by reproducing the competitive college market model used in the original study, ensuring our work is entirely reproducible. This step includes verifying the model's assumptions, computational methods, and data inputs.
2. Expanding the Study with an analysis on Canadian Universities: Specifically examine how income inequality within Canada affects university tuition, comparing these dynamics with the findings from the U.S. context. This will provide insights into how different policy environments and socio-economic conditions influence the relationship between income inequality and higher education costs.

@@

## College Tuition And Fees

```{r}
#| label: fig-stickers
#| fig-cap: College Tuition And Fees, US$(2016)
#| echo: false
#| warning: false

# Load necessary libraries
library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)
library(readr)

# Set the file path (Updated to the new path)
file_path <- "../data/analysis_data/college_tuition_fees_analysis_data.csv"

# Read the data from the CSV file (Updated function and removed sheet/range arguments)
data <- read_csv(file_path)

@@

## Comparison of Income Distribution for 1989 and 2016

```{r}
#| label: fig-log
#| fig-cap: Comparison of Income Distribution for 1989 and 2016
#| echo: false
#| warning: false
#| message: false

@@

# Define the path to your Excel file (Please update path to your own path where the data has been downloaded)
file_path <- "/Users/michaelfang/Downloads/SCF_AER_submit.xlsx"

@@

# Plot the data
ggplot(data_long, aes(x = GridLogIncome, y = EMG_ML_PDF, color = Year)) +
  geom_line() +
  labs(
       x = "Log Household Income",
       y = "Density") +
  theme_minimal() +
  scale_color_manual(values = c("2016" = "blue", "1989" = "red"))
```

@fig-log is a comparison of estimated income distributions for two different years, 1989 and 2016, represented using a probability density function on a logarithmic scale for household income.

The x-axis represents the logarithm of household income, indicating that the data is not raw income but transformed using a logarithmic function, likely to normalize the distribution and to handle the wide range of incomes more effectively. The use of logarithms in income distribution is common as it tends to make the distribution more symmetric and resemble a normal distribution, which is easier to work with statistically.

The y-axis represents the density, which is the probability per unit on the x-axis. In a probability density function, the area under the curve for a given interval represents the probability that a randomly selected household will have an income within that interval.

The increase in the variance of log income suggests that income inequality has widened from 1989 to 2016. This finding correlates with the hypothesis that growing income inequality might be one of the drivers behind the increase in college tuition. This could indicate that while the relative differences in income may have not shifted dramatically on a logarithmic scale, the actual incomes have grown, particularly at the higher end, contributing to the right skewness seen in the 2016 distribution.
